---
title: "Laws of probability"
subtitle: "Probability and Statistics"
affiliation: "Kyiv School of Economics"
format: 
  html: default
toc: true
number-sections: true
echo: false
eval: false
---

# Solving for probability of A and B

If events A and B are independent, and A has a 40% chance of happening, and event B has a 20% chance of happening, what is the probability they will both happen?

Possible answers:

- 8%
- 12%
- 20%
- 60%

# Simulating the probability of A and B

You can also use simulation to estimate the probability of two events both happening.

- Randomly simulate 100,000 flips of coin A, each of which has a 40% chance of being heads. Save this as a variable `A`.
- Randomly simulate 100,000 flips of coin B, each of which has a 20% chance of being heads. Save this as a variable `B`.
- Use the "and" operator (`&`) to combine the variables `A` and `B` to estimate the probability that both A and B are heads.

```{r}
# Simulate 100,000 flips of a coin with a 40% chance of heads
A <- rbinom(100000, 1, .4)

# Simulate 100,000 flips of a coin with a 20% chance of heads
B <- rbinom(100000, 1, .2)

# Estimate the probability both A and B are heads
mean(A & B)
```

# Simulating the probability of A, B, and C

Randomly simulate 100,000 flips of A (40% chance), B (20% chance), and C (70% chance). What fraction of the time do all three coins come up heads?

- You've already simulated A and B. Now simulate 100,000 flips of coin C, where each has a 70% chance of coming up heads.
- Use `A`, `B`, and `C` to estimate the probability that all three coins would come up heads.

```{r}
# You've already simulated 100,000 flips of coins A and B
A <- rbinom(100000, 1, .4)
B <- rbinom(100000, 1, .2)

# Simulate 100,000 flips of coin C (70% chance of heads)
C <- rbinom(100000, 1, .7)

# Estimate the probability A, B, and C are all heads
mean(A & B & C)
```

# Solving for probability of A or B

If coins A and B are independent, and A has a 60% chance of coming up heads, and event B has a 10% chance of coming up heads, what is the probability either A or B will come up heads?

- 6%
- 50%
- 64%
- 70%

# Simulating probability of A or B

In the last multiple choice exercise you found that there was a 64% chance that either coin A (60% chance) or coin B (10% chance) would come up heads. Now you'll confirm that answer using simulation.

- Use `rbinom()` to simulate 100,000 flips of coin A, each having a 60% chance of being heads.
- Use `rbinom()` to simulate 100,000 flips of coin B, each having a 10% chance of being heads.
- Use these to estimate the probability that A or B is heads.

```{r}
# Simulate 100,000 flips of a coin with a 60% chance of heads
A <- rbinom(100000, 1, .6)

# Simulate 100,000 flips of a coin with a 10% chance of heads
B <- rbinom(100000, 1, .1)

# Estimate the probability either A or B is heads
mean(A | B)
```

# Probability either variable is less than or equal to 4

Suppose X is a random Binom(10, .6) variable (10 flips of a coin with 60% chance of heads) and Y is a random Binom(10, .7) variable (10 flips of a coin with a 70% chance of heads), and they are independent.

What is the probability that either of the variables is less than or equal to 4?

- Simulate 100,000 draws from each of X (10 coins, 60% chance of heads) and Y (10 coins, 70% chance of heads) binomial variables, saving them as `X` and `Y` respectively.
- Use these simulations to estimate the probability that either X or Y is less than or equal to 4.
- Use the `pbinom()` function to calculate the exact probability that X is less than or equal to 4, then the probability that Y is less than or equal to 4.
- Combine these two exact probabilities to calculate the exact probability that either variable is less than or equal to 4.

```{r}
# Use rbinom to simulate 100,000 draws from each of X and Y
X <- rbinom(100000, 10, .6)
Y <- rbinom(100000, 10, .7)

# Estimate the probability either X or Y is <= to 4
mean(X <= 4 | Y <= 4)

# Use pbinom to calculate the probabilities separately
prob_X_less <- pbinom(4, 10, .6)
prob_Y_less <- pbinom(4, 10, .7)

# Combine these to calculate the exact probability either <= 4
prob_X_less + prob_Y_less - prob_X_less * prob_Y_less
```

# Expected value of multiplying a random variable

If X is a binomial with size 50 and p = .4, what is the expected value of 3*X?

- 1.2
- 20
- 60
- 150

# Simulating multiplying a random variable

In this exercise you'll use simulation to confirm the rule you just learned about how multiplying a random variable by a constant effects its expected value.

- Simulate 100,000 draws of X, a binomial random variable with size 20 and p = .1. Save this as `X`
- Use this simulation to estimate the expected value of X.
- Use this simulation to estimate the expected value of 5*X, as well.

```{r}
# Simulate 100,000 draws of a binomial with size 20 and p = .1
X <- rbinom(100000, 20, .1)

# Estimate the expected value of X
mean(X)

# Estimate the expected value of 5 * X
mean(5 * X)
```

# Variance of a multiplied random variable

In the last exercise you simulated X from a binomial with size 20 and p = .1 and now you'll use this same simulation to explore the variance.

- Use this simulation to estimate the variance of X.
- Estimate the variance of 5 * X

```{r}
# X is simulated from 100,000 draws of a binomial with size 20 and p = .1
X <- rbinom(100000, 20, .1)

# Estimate the variance of X
var(X)

# Estimate the variance of 5 * X
var(5 * X)
```

# Solving for the sum of two binomial variables

If X is drawn from a binomial with size 20 and p = .3, and Y from size 40 and p = .1, what is the expected value (mean) of X + Y?

- 4
- 6
- 10
- 60

# Simulating adding two binomial variables

In the last multiple choice exercise, you found the expected value of the sum of two binomials. In this problem you'll use a simulation to confirm your answer.

- Simulate 100,000 draws from X, a binomial with size 20 and p = .3, and Y, with size 40 and p = .1.
- Use this simulation to estimate the expected value of X + Y.

```{r}
# Simulate 100,000 draws of X (size 20, p = .3) and Y (size 40, p = .1)
X <- rbinom(100000, 20, .3) 
Y <- rbinom(100000, 40, .1)

# Estimate the expected value of X + Y
mean(X + Y)
```

# Simulating variance of sum of two binomial variables

In the last multiple choice exercise, you examined the expected value of the sum of two binomials. Here you'll estimate the variance.

- Use your simulation of the variables X and Y to estimate the variance of `X + Y`.
- Use your simulation to estimate the variance of `3 * X + Y`.

```{r}
# Simulation from last exercise of 100,000 draws from X and Y
X <- rbinom(100000, 20, .3) 
Y <- rbinom(100000, 40, .1)

# Find the variance of X + Y
var(X + Y)

# Find the variance of 3 * X + Y
var(3 * X + Y)
```

