---
title: "AE 07: Weights of books"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
execute: 
  echo: true
  warning: false
  eval: true
editor_options: 
  chunk_output_type: console
---

Today we'll explore the question "How do volume and weights books relate?" and "How, if at all, does that change when we take whether the book is hardback or paperback into consideration?"

# Goals

-   Build, fit, and interpret linear models with more than one predictor

-   Use categorical variables as a predictor in a model

-   Compute $R^2$ and use it to select between models

# Packages

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(tidymodels)
library(DAAG)
```

# Data

The data for this application exercise comes from the `allbacks` dataset in the **DAAG** package. The dataset has `r nrow(DAAG::allbacks)` observations and `r ncol(DAAG::allbacks)` columns. Each observation represents a book. Let's take a peek at the data:

```{r}
#| label: peek-data

allbacks
```

Note that `volume` is measured in cubic centimeters and `weight` is measured in grams. More information on the dataset can be found in the documentation for `allbacks`, with `?allbacks`.

# Single predictor

## Exercise 1

Visualize the relationship between volume (on the x-axis) and weight (on the y-axis). Overlay the line of best fit. Describe the relationship between these variables.

```{r}
#| label: plot-weight-volume

# add code here
```

> Add response here.

## Exercise 2

Fit a model predicting weight from volume for these books and save it as `weight_fit`. Display a tidy output of the model.

```{r}
#| label: fit-weight-volume

# add code here
```

## Exercise 3

Interpret the slope and the intercept in context of the data.

> Add response here.

## Exercise 4

Calculate the $R^2$ of this model and interpret it in context of the data.

```{r}
#| label: glance-weight-volume

# add code here
```

> Add response here.

# Multiple predictors

## Exercise 5

Visualize the relationship between volume (on the x-axis) and weight (on the y-axis), taking into consideration the cover type of the book. Use different colors and shapes for hardback and paperback books. Also use different colors for lines of best fit for the two types of books. In addition, add the overall line of best fit (from Exercise 1) as a gray dashed line so that you can see the difference between the lines when considering and not considering cover type.

```{r}
#| label: plot-weight-volume-cover

# add code here
```

## Exercise 6

Fit a model predicting weight from volume for these books and save it as `weight_cover_fit`. Display a tidy output of the model.

```{r}
#| label: fit-weight-volume-cover

# add code here
```

## Exercise 7

In the model output we have a variable `coverpb`. Why is only the `pb` (paperback) level of the `cover` variable shown? What happened to the `hb` (hardback) level?

> Add response here.

## Exercise 8

Interpret the slopes and the intercept in context of the data.

> Add response here.

## Exercise 9

First, guess whether the $R^2$ of this model will be greater than, less than, or the same as the previous model, or whether we can't tell. Then, calculate the $R^2$ of this model to confirm your guess, and then interpret it in context of the data.

```{r}
#| label: glance-weight-volume-cover

# add code here
```

> Add response here.

## Exercise 10

Which model is preferred for predicting book weights and why?

> Add response here.

## Exercise 11

Using the model you chose, predict the weight of a hardcover book that is 1000 cubic centimeters (that is, roughly 25 centimeters in length, 20 centimeters in width, and 2 centimeters in height/thickness).

```{r}
#| label: predict

# add code here
```