---
title: "AE 04: MINFIN scraping"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
execute: 
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

# Goals


# Packages

```{r}
#| label: load-packages

library(tidyverse)
library(rvest)
```

# Data

Let's scrape the data from the [MINFIN](https://minfin.com.ua/ua/currency/mb/archive/usd/01-02-2022/).

```{r}
#| label: load-data

url <- "https://minfin.com.ua/ua/currency/mb/archive/usd/01-02-2022/"
```


## Exercise 1

Read the content

```{r}
#| label: read-content
#| eval: false

page <- ___(url)
```

## Exercise 2

Take out the tables

```{r}
#| label: extract-tables
#| eval: false

table <- ___(page)
```

## Exercise 3

Look at the list and keep only what we are interested in. 

```{r}
#| label: clean-table
#| eval: false

table_tbl <- # your code here
```

## Exercise 4

Collect data for a certain period with a loop or a function.

In variables `custom_period` we have the custom period for which we want to collect the data.

```{r}
#| label: collect-data

custom_period <- seq(ymd("2021-02-01"), ymd("2021-02-24"), by = "1 day")
custom_period

# your code here
```

## Exercise 5

Visualize the data

```{r}
#| label: visualize-data

# your code here
```