---
title: "AE 05: Superheroes"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
execute: 
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

# Packages

```{r}
#| label: packages
#| message: false
library(tidyverse)
```

# Data

Load the data on [superheroes](https://aranaur.rbind.io/datasets/heroes/).

## Variables

* `name` --- superhero name
* `Gender` --- gender of the superhero
* `Eye color` --- eye colors
* `Race` --- race of the superhero
* `Hair color` --- hair colors
* `Height` --- height of the superhero
* `Publisher` --- publisher
* `Skin color` --- skin color
* `Alignment` --- moral alignment of the superhero
* `Weight` --- weight of the superhero

## Exercise 1

Imagine that our superheroes from the heroes dataset are a sample from the general population of all written and unwritten superheroes. Perform a one-sample $t$-test for the weight of the superheroes and the number `100`, which is the estimated average weight in the general population of all superheroes. Interpret the result.

```{r}
#| label: exercise1
#| eval: false

t.test(___, mu = 100)
```

## Exercise 2

Perform a one-sample $t$-test for the height of superheroes, and the number is 185, which is the estimated average height in the general population of all superheroes. Interpret the result.

```{r}
#| label: exercise2
#| eval: false

t.test(___, mu = 185)
```

## Exercise 3

Perform an independent $t$-test to compare the masses of superheroes with black and white eyes.

```{r}
#| label: exercise3
#| eval: false

t.test(___ ~ ___, data = ___)
```

