---
title: "AE 06: Apple and Microsoft stock prices"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
execute: 
  echo: true
  warning: false
  eval: false
editor_options: 
  chunk_output_type: console
---

Today we'll explore the question "How do stock prices of Apple and Microsoft relate to each other?"

# Goals

-   Understand the grammar of linear modeling, including $y$, $x$, $b$, $e$ fitted estimates and residuals

-   Add linear regression plots to your 2D graphs

-   Write a simple linear regression model mathematically

-   Fit the model in R.

# Packages

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(tidyquant)
```

# Data

The data for this application exercise is gathering with the [**tidyquant**](https://business-science.github.io/tidyquant/) R package. It features Apple and Microsoft stock prices from January 1st 2020 to December 31st 2020.

First, let's load the data:

```{r}
#| label: load-data
#| message: false

stocks <- tq_get(c("AAPL", "MSFT"), from = "2020-01-01", to = "2021-01-31") %>% 
  select(date, symbol, open) %>% 
  pivot_wider(names_from = symbol, values_from = open)
```

# Models and residuals

## Exercise 1

At first, you might be tempted to minimize $\sum_i e_i$, the sum of all residuals, but this is problematic. Why? Can you come up with a better solution (other than the one listed below)?

*Add response here.*

## Minimize the sum of squared residuals

In practice, we minimize the **sum of squared residuals**:

$$
\sum_i e_i^2
$$

Note, this is the same as

$$
\sum_i (y_i - \hat{y}_i)^2
$$

## Exercise 2

Check out an interactive visualization of "least squares regression" [here](https://seeing-theory.brown.edu/regression-analysis/index.html#section1). Click on `I` and drag the points around to get started. Describe what you see.

*Add response here*

# Plotting the least squares regression line

Plotting the OLS regression line, that is, the line that minimizes the sum of square residuals takes one new geom. Simply add

```         
geom_smooth(method = "lm", se = FALSE)
```

to your plot.

`method = "lm"` says to draw a line according to a "linear model" and `se = FALSE` turns off standard error bars. You can try without these options for comparison.

Optionally, you can change the color of the line, e.g.

```         
geom_smooth(method = "lm", se = FALSE, color = "#e64173")
```

## Exercise 4

In the slides we fit a model with slope 0.5 and intercept -5. The code for layering the line that represents the model over your data is given below. Add `geom_smooth()` as described above with `color = "steelblue"` to see how close your line is.

```{r}
#| label: how-close

ggplot(stocks, aes(x = MSFT, y = AAPL)) +
  geom_point() + 
  geom_abline(slope = 0.5, intercept = -5) +
  # your code here
  labs(
    x = "MSFT Open", 
    y = "AAPL Open", 
    title = "Open prices of MSFT and AAPL",
    subtitle = "January 2020"
  )
```

# Finding $\hat{\beta}$

To **fit the model** in R, i.e. to "find $\hat{\beta}$", use the code below as a template:

```{r}
#| eval: false       
model_fit <- lm(MSFT ~ AAPL, data = stocks)
```

-   `lm()` stands for "linear model"
-   `MSFT ~ AAPL` means "MSFT is the dependent variable, AAPL is the independent variable"

Running the code above, you will get a model object `model_fit` that contains all the information about the model.

To access the information, you can run, e.g.

```{r}
summary(model_fit)
```

Let's try it out.

## Exercise 5

Find the least squares line $y = \hat{\beta_0} + \hat{\beta_1} x$ for January 2020, where $x$ is Microsoft's opening price and $y$ is Apple's opening price. Display a tidy summary of your model.

```{r}
#| label: fit-MSFT-AAPL-January

# code here
```

## Exercise 6

Re-write the fitted equation replacing $\hat{\beta}_0$ and $\hat{\beta}_1$ with the estimates from the model you fit in the previous exercise.

$$
\text{[add equation here]}
$$

## Exercise 7

Interpret $\hat{\beta}_0$ and $\hat{\beta}_1$ in context of the data.

*Add response here*

## Bonus exercise

Say Microsoft opens at 166 dollars. What do I predict the opening price of AAPL to be?

```{r}
#| label: predict

# add code here
```