---
title: "AE 02: Exploring Flint's water data"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
execute: 
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

# Goals

By the end of this application exercise you will

-   meet the computational toolkit for the course
-   practice using `glimpse()`, `names()`, `nrow()`, `ncol()`, `count()`
-   define and compute various statistics

We will do this using water [lead content data from Flint, MI](https://en.wikipedia.org/wiki/Flint_water_crisis). The following paragraph will be useful in evaluating the lead amount values we'll see in the dataset.

> While there is no completely safe amount of lead consumption, the limit allowed by the Lead and Copper Rule (LCR) of 1991 is 15 parts per billion (ppb). If this is exceeded in more than 10% of homes tested (or if the 90th percentile value of the total sample is above 15 ppb), action is required. And to make sure problems are caught, sampling for lead in water is supposed to target the "worst-case" homes -- those in areas served by lead pipes.

## Load packages

We'll use the tidyverse package for analysis, which offers functionality for data import, wrangling, visualization, and more.

```{r}
#| label: load-packages
#| warning: false

library(tidyverse)
```

## Load data

Load the [Flint water data set](https://aranaur.rbind.io/datasets/flint/). Name the data frame `flint`.

The `read_csv()` function can be used for reading CSV (comma separated values) files. The file we're reading is called `flint` with the suffix (`.csv`) which indicates its file type.

```{r}
#| label: load-data
#| eval: false

flint <- # add code here
```

## Data dictionary

The following variables are in the `flint` data frame:

-   `id`: sample ID number (identifies the home)
-   `zip`: ZIP code in Flint of the sample's location
-   `ward`: ward in Flint of the sample's location
-   `draw`: which time point the water was sampled from
-   `lead`: lead content in parts per billion (ppb)

## Populations and samples

We want to learn about the **population** using a **sample**.

In the case we want to learn about the lead content in all of Flint, MI homes but only have available water readings from a sample of homes (our data set).

**Exercise 1:** Look at the data, how many observations are there? How many variables?

```{r}
#| label: obs-vars
#| eval: false

# add code here 
```

## Frequencies

Let's `count()` to find the number of different time points water was sampled with the `count()` function.

-   The first argument is `flint`: the data frame
-   The second argument is `draw`: the variable

```{r}
#| label: count-draws
#| eval: false

count(flint, draw)
```

We can achieve the same result with the following "piped" operation as well.

-   The first line is `flint`: the data frame

-   Then the pipe operator, read as "and then", which places what comes before it as the first argument of what comes after it

-   The second line is `count(draw)`

```{r}
#| label: count-draws-piped
#| eval: false

flint %>%
  count(draw)
```

We can use a similar approach to fund out how many unique homes are in the data set:

```{r}
#| label: count-homes
#| eval: false

flint %>%
  count(id)
```

**Exercise 2:** How many samples were taken from each zip code?

```{r}
#| label: count-by-zip

# add code here
```

**Exercise 3:** Which ZIP code had the most samples drawn? Hint: See the help for `count`.

```{r}
#| label: count-by-zip-ordered

# add code here
```

# Summarization

What is a statistic? It's **any mathematical function of the data**. Sometimes, a statistic is referred to as "sample statistic" or "summary statistic", since you compute it from a **sample** (the observed data) and not the entire **population**.

## Measures of central tendency

-   mean
-   median
-   mode

... and any other arbitrary function of the data you can come up with!

**Exercise 4:** Compute each of these statistics for `lead` ppb.

```{r}
#| label: statistics

# add code here
```


# References

1.  Langkjaer-Bain, R. (2017). The murky tale of Flint's deceptive water data. Significance, 14: 16-21. doi: <https://doi.org/10.1111/j.1740-9713.2017.01016.x>.

2.  Kelsey J. Pieper, et. al. Evaluating Water Lead Levels During the Flint Water Crisis. Edwards Environmental Science & Technology 2018 52 (15), 8124-8132 doi: <https://doi.org/10.1021/acs.est.8b00791>.
