---
title: "Lab 01"
subtitle: "Descriptive Statistics"
author: "Ihor Miroshnychenko"
institute: Kyiv School of Economics
format:
  html: 
    highlight-style: github
    theme: [default, custom.scss]
    toc: true
    toc-depth: 2
execute: 
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

The goal of this lab is to effectively visualize numerical and categorical data.

For all visualizations you create, b sure to include informative titles for the plot, axes, and legend!

# Packages

In this lab we will work with the tidyverse packages, which is a collection of packages for doing data analysis in a "tidy" way.

```{r}
#| label: load-packages
#| message: false

library(tidyverse) 
```

# NC Courage

Today, we will be working with data from the [first three full seasons of the NC Courage](https://aranaur.rbind.io/datasets/courage/), a highly successful National Women's Soccer League (NWSL) team located near Duke in Cary, NC. The Courage moved to the Triangle from Western New York in 2017 and had three epic seasons in NC, culminating in winning the championship game that was held at their stadium in Cary in 2019! Data for this lab was sourced from the [`nwslR` package on Github](https://github.com/adror1/nwslR).

The variables in the dataset are as follows:

| Variable      | Description                                        |
|---------------|----------------------------------------------------|
| `game_id`     | An unique ID for the game                          |
| `game_date`   | Game date                                          |
| `game_number` | Game number                                        |
| `home_team`   | Name of the home team, abbreviated                 |
| `away_team`   | Name of the away team, abbreviated                 |
| `opponent`    | The team NC Courage played against                 |
| `home_pts`    | Number of points by the home team                  |
| `away_pts`    | Number of points by the away team                  |
| `result`      | Result of the game for NC Courage (win, loss, tie) |
| `season`      | Season (2017, 2018, or 2019)                       |

## Exercise 1

::: border
Create a bar plot of the `result`s of games for NC Courage. Additionally, calculate the numbers of wins, losses, and ties. Write a one sentence narrative for your findings.
:::

::: {.callout-tip}
`result` is a categorical variable, so use a bar plot for the visualization and the `count()` function for calculating the frequencies of *levels* of this variable.
:::

## Exercise 2

::: border
Create a new variable indicating whether the game was played at home or away for NC Courage. This variable should be called `home_courage` and take the value "home" if NC Courage is the home team and "away" if NC Courage is the away team.

Then, calculate the number of home and away games, and write a one sentence narrative for your findings.
:::

## Exercise 3

::: border
First, create a visualization that displays the relationship between `home_courage` and `result`. Then, calculate the proportions of home and away games that the Courage won. Based on these, do your findings suggest a home-field advantage? Why or why not?
:::

So far we have focused on whether the game was at home or away and whether the Courage won. Next, we dive deeper and focus on the number of points the Courage wins by, at home and away.

## Exercise 4

::: border
How many points do the Courage typically win by (on average)?
:::

::: {.callout-tip}
We are only interested in games the Courage wins, therefore we should `filter()` for those games first.
:::

## Exercise 5

::: border
How many points do NC Courage score when they win (on average)? Note this is **different than how many points** they "win by". How many points do the Courage score when they lose on average?
:::

To calculate this we first need to determine how many points NC Courage scored in every game. We can use `if_else()` logic to find this value for each game, and store it in a new column, `courage_pts`.

## Exercise 6

::: border
Next we'll investigate visually whether or not NC Courage has a home-field advantage.

Mutate the `courage` data frame to create two new variables:

-   `total_pts`: Sum of points scored by both teams, i.e. `home_pts + away_pts`.
-   `opponent_pts`: Points scored by the opposing team, i.e., `total_pts - courage_pts`.

Save the resulting data frame as `courage` again and print the three points columns (`total_pts`, `opponent_pts`, `courage_pts`) to screen.
:::

## Exercise 7

::: border
Create a scatter plot:

-   `opponent_pts` (y) vs. `courage_pts` (x)
-   Color the scatter plot by whether NC Courage are home or away.
-   Represent the data with "jittered" points wth `geom_jitter()`.
-   Overlay a $y = x$ line with `geom_abline()`.
-   Faceted by season.

What does the line represent? What does it mean for a point to fall above the line? Below the line?
:::

```{r}
#| eval: false

ggplot(courage, aes(x = ___, y = ___, color = ___)) + 
  geom_jitter(width = 0.1, height = 0.1) + 
  geom_abline(slope = 1, intercept = 0) +
  facet_wrap(~ ___) +
  labs(
    x = "___", 
    y = "___", 
    title = "___", 
    color = "___"
  )
```

## Exercise 8

This exercise is a look at where we're headed...

::: border
If we want to formally test whether the Courage have a home-field advantage, then we must first define what this means! In your own words, what do you think a home-field advantage means? Then, now that you've defined what it means to have a home field advantage, define what it means to **not** have a home-field advantage.
:::

::: callout-note
While there is a right answer, this part is graded for completion, so don't worry too much about answering this in *exactly* the right way. Although graded for completion, your response must make sense to receive full points.
:::

# Grading

- **Submission of Lab Report:** A complete and accurate submission of the lab assignment in the required format [(1 point)]{.hi}.
- **Understanding of Code and Correct Use of Methods:** Evaluation of your understanding of the code you submit and the appropriateness of the methods used in your solution [(2 point)]{.hi}.
- **Verbal Defense of the Lab:** An oral explanation of your solution to verify your understanding of the concepts and methods used. You may be asked to explain the logic of your code, the interpretation of results, or to make minor modifications to your solution during the defense [(2 point)]{.hi}.
